---
title: "Smoking and the risk of disease"
author: "Gideon Potgieter"
date: "2/20/23"
output: html_document
---

##### Comments from Dr. Taub: 

* Good job!  
* You can see any changes/comments made directly in your document by searching for "MAT" in the search box in RStudio cloud.
* You should include captions for the figures

**Comments on Question 1**:
* For Q1, your graph, analysis, and results are very nicely done! Some small comments for improvement are below. 
* While you provide a good description of what you see in the plot, I would like to see more quantitative support for your statements, using actual numbers from the visualization.
* Your interpretation of your odds ratio for both models is correct -- to make it a nice numerate sentence we can just stick the confidence interval in parentheses at the end rather than giving it its own sentence.  So: To find out if the difference is statistically significant, a logistic regression model was created that shows that the odds of having MSCD for smokers was 1.7 times the odds of having MSCD for non-smokers who had the same socioeconomic status, sex, and education level (p < 0.05, 95% CI (1.29, 2.27)). 

**Comments on Question 2**:
* Overall, there is too much raw R output and code visible. You don't need to describe the recoding of the variables, except to clearly state what the outcome is defined as.
* Nice graphs here, effectively showing the relative risks for the groups of interest. Your titles are getting truncated; you can alleviate this by including a newline character "\n" to break the title across multiple lines. However, in your write-up describing what they show, you should include specific numbers from the visualizations to support your statements.
* Please include a caption on the tables indicating what variables are included in your propensity score model, so that it is clear without reading the full text what variables are being controlled for in the analysis. You do not need to include table entries for the propensity score quintiles, just the variables relevant to the question (main effects and interaction terms)
* Your interpretation of the model results is correct and overall concise, but I would suggest including the numerical information about statistical significance in a parenthetical statement rather than its own sentence.
* Why did you choose to use propensity scores here in contrast to your analysis for Question 1? And to control for age here, where you did not control for it above? It is OK to do it this way, but might be worth thinking about or addressing.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

MAT: You can remove this line in your final report
**Due Monday, February 20 @ midnight through Github**

```{r load_libraries, echo=FALSE, include=FALSE}
#load any of the libraries you will use in this code chunk
library(tidyverse)
```

```{r read_data, echo=FALSE, include=FALSE}
# Read in the data
nmes_data <- read_csv("nmesUNPROC.csv")

library(broom) 
library(kableExtra) 
library(gtsummary)
```

```{r recode_data, echo=FALSE, include=FALSE}
# Place any code that you used to recode the data here
nmes_data <- nmes_data %>%
  mutate(eversmk = factor(eversmk, levels = c("0", "1"), labels = c("Never smoker", "Ever smoker")),
         lc5 = factor(lc5, levels = c("0", "1"), labels = c("No LC", "LC")),
         chd5 = factor(chd5, levels = c("0", "1"), labels = c("No CHD", "CHD")),
         female = factor(female, levels= c("0", "1"), labels = c("Male", "Female")),
         current = factor(current, levels= c("0", "1"), labels = c("Not current smoker", "Current smoker")),
         former = factor(former, levels= c("0", "1"), labels = c("Not former smoker", "Former smoker")),
         beltuse = factor(beltuse, levels= c("1", "2", "3"), labels = c("Rare", "Some", "Almost always")),
         educate = factor(educate, levels= c("1", "2", "3", "4"), labels = c("College grad", "Some college", "HS grad", "Other")),
         marital = factor(marital, levels= c("1", "2", "3", "4", "5"), labels = c("Married", "Widowed", "Divorced", "Separated", "Never married")),
         poor = factor(poor, levels= c("0", "1"), labels = c("Not poor", "Poor"))
         )

nmes_data <- nmes_data %>%
  mutate(disease = factor(lc5 == "LC" | chd5 == "CHD", 
                          levels=c(FALSE,TRUE), 
                          labels=c("No MSCD", "MSCD")))
```

**Question 1.1: How does the risk of disease compare for smokers and otherwise similar non-smokers?**

```{r q1_display, echo=FALSE}
### put code to create data display here
Table_1 <- nmes_data %>%
  count(eversmk, female, poor, disease) %>%
  group_by(female, poor, eversmk) %>%
  mutate(prop = n/sum(n)) %>%
   mutate(percent =round(n/sum(n)*100,1))

Table_1 %>%
  filter(disease == "MSCD") %>%
  ggplot() +
  geom_bar(mapping = aes(x = eversmk, y = percent, fill = female), stat = "identity") + facet_grid(female~poor) +
   geom_text(aes(x = eversmk, y = percent, label = percent)) +
  labs(y= "Risk of MSCD (%)",
       x= "",
       title= "Risk of MSCD, comparing smokers to non-smokers grouped by income and sex", 
       fill = "Sex")
```

```{r q1_analysis, echo=FALSE}
### put code to fit statistical model and produce table of results here
### you could use either your logistic regression model or your propensity scores model
### or you could include both and compare the results

model1 <- glm(disease ~ eversmk + poor + female + educate, family=binomial(link="logit"), data=nmes_data)

model1 %>%
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.96) %>%
  filter(term != "(Intercept)") %>%
  mutate(conf.int = paste0("(", round(conf.low,2), ", ", round(conf.high,2), ")")) %>%
  select(Term = term, OR = estimate, `p-value` = p.value, `95% CI` = conf.int) %>%
  kable(digits = 3, format = "markdown")
```

To find out how the risk of disease compares for smokers and otherwise similar non-smokers, a bar plot was created that shows the risk of MSCD as a proportion grouped by socioeconomic status and sex. In all four subgroups, people who have ever smoked were more likely to have MSCD, which supports the hypothesis that the risk of disease is greater for smokers than otherwise similar non-smokers. To find out if the difference is statistically significant, a logistic regression model was created that shows that the odds of having MSCD for smokers was 1.7 times the odds of having MSCD for non-smokers who had the same socioeconomic status, sex, and education level. The p-value was 0.000, meaning it is statistically significant, and the 95% confidence interval was (1.29, 2.27). Therefore, the risk of disease is higher for smokers than for otherwise similar non-smokers.


**Question 1.2: Does the contribution of smoking to the risk of disease vary by sex or socio-economic status (SES)?**

```{r q2_display, echo=FALSE}
### put code to create data display here
my_table <- nmes_data %>%
  count(poor, female, eversmk, disease) %>%
  group_by(poor, female, eversmk) %>%
  mutate(prop = n/sum(n)) %>%
  filter(disease == "MSCD") %>%
  ungroup() %>%
  group_by(poor, female) %>%
  mutate(rel.risk = round(prop/prop[eversmk == "Never smoker"],1))

my_table %>%
  filter(eversmk == "Ever smoker") %>%
  ggplot() +
  geom_bar(aes(x = female, y = rel.risk, fill = female), stat = "identity") +
  facet_wrap(~ poor) + 
  geom_text(aes(x = female, y = rel.risk, label = rel.risk)) +
  labs(y= "Relative Risk of MSCD",
       x= "",
       title= "Relative risk of MSCD, comparing male smokers to female smokers, holding socioeconomic status constant", 
       fill = "Sex")

my_table %>%
  filter(eversmk == "Ever smoker") %>%
  ggplot() +
  geom_bar(aes(x = poor, y = rel.risk, fill = poor), stat = "identity") +
  facet_wrap(~ female) + 
  geom_text(aes(x = poor, y = rel.risk, label = rel.risk)) +
  labs(y= "Relative Risk of MSCD",
       x= "",
       title= "Relative risk of MSCD, comparing poor smokers to not poor smokers, holding sex constant", 
       fill = "poor")
```

```{r q2_analysis, echo=FALSE}
### put code to fit statistical model and produce table of results here
#prop model for female:

my_data <- nmes_data

prop_model <- glm(eversmk ~ age + female, family = binomial(link="logit"), data=my_data, na.action = na.exclude)

my_data <- my_data %>%
  mutate(ps = predict(prop_model, type = "response"))

ps_quintiles <- quantile(my_data$ps, probs=c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm=TRUE)

my_data <- my_data %>%
  mutate(ps_strata = cut(ps, breaks=ps_quintiles, include.lowest=TRUE))

model_ps_strata <- glm(disease ~ eversmk*female + ps_strata, family = binomial(link="logit"), data=my_data)
tidy(model_ps_strata)

tidy(model_ps_strata, exponentiate = TRUE, conf.int = TRUE )

model_ps_strata %>%
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.96) %>%
  filter(term != "(Intercept)") %>%
  mutate(conf.int = paste0("(", round(conf.low,2), ", ", round(conf.high,2), ")")) %>%
  select(Term = term, OR = estimate, `p-value` = p.value, `95% CI` = conf.int) %>%
  kable(digits = 3, format = "markdown")

#prop model for SES:

my_data <- nmes_data

prop_model <- glm(eversmk ~ age + poor, family = binomial(link="logit"), data=my_data, na.action = na.exclude)

my_data <- my_data %>%
  mutate(ps = predict(prop_model, type = "response"))

ps_quintiles <- quantile(my_data$ps, probs=c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm=TRUE)

my_data <- my_data %>%
  mutate(ps_strata = cut(ps, breaks=ps_quintiles, include.lowest=TRUE))

model_ps_strata <- glm(disease ~ eversmk*poor + ps_strata, family = binomial(link="logit"), data=my_data)
tidy(model_ps_strata)

tidy(model_ps_strata, exponentiate = TRUE, conf.int = TRUE )

model_ps_strata %>%
  tidy(exponentiate = TRUE, conf.int = TRUE, conf.level = 0.96) %>%
  filter(term != "(Intercept)") %>%
  mutate(conf.int = paste0("(", round(conf.low,2), ", ", round(conf.high,2), ")")) %>%
  select(Term = term, OR = estimate, `p-value` = p.value, `95% CI` = conf.int) %>%
  kable(digits = 3, format = "markdown")
```

The first display shows the relative risk of MSCD comparing male smokers to female smokers and holding socioeconomic status constant, while the second display shows the relative risk of MSCD comparing poor smokers to not poor smokers holding sex constant.  The relative risk was higher for male smokers than female smokers in both socioeconomic groups, and the relative risk was higher for poor smokers than not poor smokers in both sexes. The first multivariable logistic regression model shows that the odds ratio of disease for females is 13% lower than the odds ratio of disease for males of the same age, but this result is not statistically significant (p-value = 0.6). The second multivariable logistic regression model shows that the odds ratio of disease for poor people is 79% higher than the odds ratio for not poor people of the same age. This result is statistically significant (p-value = 0.033), so we know that socioeconomic status affects the contribution of smoking to the risk of disease.

```{r submission_instructions, echo=FALSE, include=FALSE}
#When you are ready to submit your assignment, do ALL of the following:

#* First, knit your .Rmd file to see if it will compile.  If it doesn't knit, you can still follow the steps below to submit your assignment, but please try to resolve the knitting issues before you submit.  You can reach out to use at phbiostats@jhu.edu for help!

#* Next you need to **commit** the changes you've made to the document.  Click the colorful Git button at the top of the RStudio window and select "Commit" from the menu.

#* In the window that opens, **stage** your changes by clicking the check boxes next to the `Asst1-4.Rmd` file.

#* In the "Commit message" box, type a short description of what changes you've made, something like: `Completed assignment`

#* Click the "Commit" button on the bottom right.

#* You'll see a window that will tell you about the changes that you've made.  Click "Close" to close the window.  You've successfully committed! You can close this commit window now.

#* After committing, you must **push** your changes to the repository on Github.  Do this by clicking the colorful Git button again and select "Push Branch".  

#* Again, you'll see a window open that tells you your changes have been pushed!

#* If you want, you can look at your repository on [Github.com](https://github.com/) and should be able to see your changes there!  

#* You've successfully submitted your assignment :)
```

